Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-11-10 paper-new/concrete-term-model.tex
small edits

Consider the functor $F(X) = \mathbb{N} + {\mho} + X$. By the definition of the
guarded lift, we have that
% 
$\li^k \mathbb{N} 
  \cong \mathbb{N} + {\mho} + \later^k (\li \mathbb{N}) 
  \cong F(\later \li^k \mathbb{N})$.
%
Furthermore, it can be shown that the functor $F$ satisfies the technical
condition in the statement of the above-mentioned theorem. Thus, the theorem
implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a more familiar coninductive type.
%
Given a type $Y$, Capretta's \emph{delay monad} \cite{lmcs:2265},
which we write as $\text{Delay}(Y)$, is defined as the coinductive type
generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(X)$.
We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
$F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
$\delay(\mathbb{N} + {\mho})$. Denote this isomorphism by $\alpha$.

Given $d : \delay(\mathbb{N} + {\mho})$, we define a notion of termination in
$i$ steps, written $d \da^i$. We define $d \da^i$ inductively by the rules
$(\tnow\, n_?) \da^0$ and $(\tlater\, d) \da^{i+1}$ if $d \da^i$ (here, $n_? :
\mathbb{N} + {\mho}$). From this definition, it is clear that if there is an $i$
such that $d \da^i$, then this $i$ is unique. 
% Furthermore, from the proof that
% $d$ terminates in $i$ steps, we can extract the result with which $d$
% terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
Using this definition of termination, we define a partial function 
%
$f \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
%
by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we
define $f(d) := n_?$ if $d \da^i n_?$ for some (necessarily unique) $i$, and $f$
is undefined otherwise.

>>>>

Consider the functor $F(X) = \mathbb{N} + {\mho} + X$. By the definition of the
guarded lift, we have that
% 
$\li^k \mathbb{N} 
  \cong \mathbb{N} + {\mho} + \later^k (\li \mathbb{N}) 
  \cong F(\later \li^k \mathbb{N})$.
%
Furthermore, it can be shown that the functor $F$ satisfies the technical
condition in the statement of the above-mentioned theorem. Thus, the theorem
implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a more familiar coninductive type.
%
Given a type $Y$, Capretta's \emph{delay monad} \cite{lmcs:2265},
which we write as $\text{Delay}(Y)$, is defined as the coinductive type
generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(X)$.
We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
$F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
$\delay(\mathbb{N} + {\mho})$. We write $\alpha$ to denote this isomorphism.

Given $d : \delay(\mathbb{N} + {\mho})$, we define a notion of termination in
$i$ steps with result $n_? \in \mathbb{N} + {\mho}$, written $d \da^i n_?$.
We define $d \da^i n_?$ inductively by the rules
$(\tnow\, n_?) \da^0 n_?$ and $(\tlater\, d) \da^{i+1} n_?$ if $d \da^i n_?$.
From this definition, it is clear that if there is an $i$
such that $d \da^i n_?$, then this $i$ is unique. 
% Furthermore, from the proof that
% $d$ terminates in $i$ steps, we can extract the result with which $d$
% terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
Using this definition of termination, we define a partial function 
%
$f \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
%
by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we
define $f(d) := n_?$ if $d \da^i n_?$ for some (necessarily unique) $i$, and $f$
is undefined otherwise.

### 2024-11-10 paper-new/intro.tex
add citation for Clocked Cubical Type Theory in the intro
small edits

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
However, an approach based on classical domain theory has fundamental
limitations: domain theory appears incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references \cite{Birkedal-Stovring-Thamsborg-2009}, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a denotational approach that can
scale up to these advanced features, and so we must abandon classical
domain theory as the foundation in order to make progress. In this
work, we provide first steps towards this goal by adapting work on a
simple gradually typed lambda calculus to \emph{guarded domain
theory}. Our goal is for this model to scale to dynamic type tag
generation and higher-order references in future work.

\emph{Guarded} domain theory is the main denotational alternative to
classical domain theory that can successfully model these advanced
features. While classical domain theory is based on modeling types as
ordered sets with certain joins, guarded domain theory is based on an
entirely different foundations, sometimes (ultra)metric spaces but
more commonly as ``step-indexed sets'', i.e., objects in the
\emph{topos of trees}
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
category theoretic terminology, these are presheaves on the poset of
natural numbers.)  We think of a $\mathbb{N}$-indexed set as an
infinite sequence of increasingly precise approximations to the true
type being modeled.
%
Key to guarded domain theory is that there is an operator
$\triangleright$ on step-indexed sets called ``later''. In terms of
sequences of approximations, the later operator delays the
approximation by one step. Then the crucial axiom of guarded domain
theory is that any guarded domain equation $X \cong F(\triangleright
X)$ has a unique solution. This allows guarded domain theory to model
essentially \emph{any} recursive concept, with the caveat that the
recursion is \emph{guarded} by a use of the later operator.

@@

% The move to synthetic setting is useful, but now you're stuck in the guarded setting
We have to take care when defining such a big-step semantics in
guarded type theory, as a naive definition of termination is
incorrect: we can prove using guarded recursion that an infinite loop
terminates! This is because the definition of termination inside
guarded logic has a different meaning, essentially meaning that for
every finite number $n$, either the program terminates in $< n$ steps
or it takes $n+1$ steps, which is equivalent to a statement of
progress rather than termination. For this reason, we need a kind of
``escape hatch'' in our ambient type theory to leave the guarded
setting and make definitions that are interpreted as constructions in
ordinary mathematics. We incorporate this by using
\emph{clock-quantification}\max{Eric should provide a citation for Clocked cubical here since that's close to our actual approach} \cite{atkey-mcbride2013} in our type
theory. In particular, the guarded type theory in which we work has a
notion of clocks that index all of our guarded constructions. This is
discussed in more detail in Section \ref{sec:big-step-term-semantics}.

>>>>

The only prior denotational semantics for gradual typing was given by
New and Licata and is based on a classical Scott-style \emph{domain
theory} \cite{new-licata18}. The fundamental idea is to equip
$\omega$-CPOs with an additional ``error ordering'' $\ltdyn$ which
models the graduality ordering, and for casts to arise from
\emph{embedding-projection pairs}. Then the graduality property
follows as long as all language constructs can be interpreted using
constructions that are monotone with respect to the error ordering.
%
This framework has the benefit of being compositional, and was
expressive enough to be extended to model dependently typed gradual
typing \cite{gradualizing-cic}.
%
However, an approach based on classical domain theory has fundamental
limitations: domain theory appears incapable of modeling certain perversely
recursive features of programming languages such as dynamic type tag
generation and higher-order references \cite{Birkedal-Stovring-Thamsborg-2009}, which are commonplace in
real-world gradually typed systems as well as gradual calculi.
%
Our long-term goal is to develop a denotational approach that can
scale up to these advanced features, and so we must abandon classical
domain theory as the foundation in order to make progress. In this
work, we provide first steps towards this goal by adapting work on a
simple gradually typed lambda calculus to the setting of \emph{guarded domain
theory}, a task that already involves nontrivial issues.
Our goal is for this model to scale to dynamic type tag
generation and higher-order references in future work.

Guarded domain theory is the main denotational alternative to
classical domain theory that can successfully model the advanced
features we aim to support. While classical domain theory is based on modeling types as
ordered sets with certain joins, guarded domain theory is based on an
entirely different foundations, sometimes (ultra)metric spaces but
more commonly as ``step-indexed sets'', i.e., objects in the
\emph{topos of trees}
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
category theoretic terminology, these are presheaves on the poset of
natural numbers.)  We think of a $\mathbb{N}$-indexed set as an
infinite sequence of increasingly precise approximations to the true
type being modeled.
%
Key to guarded domain theory is that there is an operator
$\triangleright$ on step-indexed sets called ``later''. In terms of
sequences of approximations, the later operator delays the
approximation by one step. Then the crucial axiom of guarded domain
theory is that any guarded domain equation $X \cong F(\triangleright
X)$ has a unique solution. This allows guarded domain theory to model
essentially \emph{any} recursive concept, with the caveat that the
recursion is \emph{guarded} by a use of the later operator.

@@

% The move to synthetic setting is useful, but now you're stuck in the guarded setting
We have to take care when defining such a big-step semantics in
guarded type theory, as a naive definition of termination is
incorrect: we can prove using guarded recursion that an infinite loop
terminates! This is because the definition of termination inside
guarded logic has a different meaning, essentially meaning that for
every finite number $n$, either the program terminates in $< n$ steps
or it takes $n+1$ steps, which is equivalent to a statement of
progress rather than termination. For this reason, we need a kind of
``escape hatch'' in our ambient type theory to leave the guarded
setting and make definitions that are interpreted as constructions in
ordinary mathematics. We incorporate this by using
\emph{clock-quantification} \cite{atkey-mcbride2013, kristensen-mogelberg-vezzosi2022}
% \max{Eric should provide a citation for Clocked cubical here since that's close to our actual approach}
in our type theory. In particular, the guarded type theory in which we work has
a notion of clocks that index all of our guarded constructions. This is
discussed in more detail in Section \ref{sec:big-step-term-semantics}.
